Demographic Species Distribution Models

### 2019-10-16 manuscript/appendix_a.Rmd
tidy up appendix; add regression model bit to main ms

This appendix provides notation and, where available, computer code for R packages greta and greta.dynamics [@goldingGretaDynamicsModelling2019; @goldingGretaSimpleScalable2019] to implement the DSDMs proposed in the manuscript *Demographic Species Distribution Models* by Nick Golding.

Note that statistical notation is written in the standard way, with the first line giving the sampling distribution, and subsequent lines defining the variables on whhich they depend, however the code uses the reverse order since in greta those variables must be defined bfore they can be used.

@@

For ease of reference, I repeat the equations and interpretation for the basic presence-absence DSDM presented in the main manuscript. It presents the model in its three parts, each of which can be modified independently. Specification of priors for parameters in the model are not discussed here. See [@goldingGretaSimpleScalable2019] and the website [greta-stats.org](https://greta-stats.org) for general guidance on how to build models with greta.

@@

Given vectors `Lambda` and `y` each with length `n`, the number of sites.

## Abundance relationship

The abundance relationship relates the transition matrix $\mathbf{A}_i$ of the population at site $i$ to the relative abundance $\Lambda_i$. The intrinsic growth rate $\lambda_i$ of the population is used to compute the theoretical carrying capacity $K_i$ according to the Beverton-Holt model, with competition effect parameter $\alpha$. This theoretical carrying capacity is negative when $\lambda_i < 1$, so it is rectified by the softplus function to always yield a positive abundance $\Lambda_i$, which is close to zero when $\lambda_i < 1$, and approximately equal to $K_i$ when $\lambda_i > 1$.

@@

Given a 3D array `A` with dimensions `n` x `m` x `m`, where `n` is the number of sites, and `m` the number of stages in the population model (dimension of the matrices $\mathbf{A}_i$), and `alpha` is a scalar. When used with its default settings the function `greta.dynamics::iterate_matrix()` returns, among other quantities, a vector of intrinsic growth rates of the matrices in `A` computed by repeated iteration of the matrix

## Transition matrix and vital rate regressions

@@

Given an `n` x `k` matrix `X`, length `k` vectors `beta_S` and `beta_F`, and a scalar `gamma`, where `n` is the number of sites and `k` the number of columns (covariates plus an intercept) of the design matrix.

# Alternative observation models

@@

# Alternative abundance relationships

@@

analysis <- iterate_dynamic_matrix(
  matrix_function = A_t_function, 
  base_log_F = X %*% beta_F,
  base_logit_S = X %*% beta_S,
  alpha_F = alpha_F,
  alpha_S = alpha_S,
  gamma = gamma
)

Lambda <- analysis$stable_population
```
Given additional scalar parameters `alpha_F` and `alpha_S`.

# Temporally-varying abundance

@@

## Long-term environmental variation

@@

## Short term environmental variation

Where the time period between observations is sufficiently short that the assumptions above are not met, or when the ecological question being answered requires it, temporal fluctuation in vital rates can instead be modelled by explicitly iterating the population through successive timesteps, at each time step using the corresponding transition matrix, as in the density-dependent vital rate model above. Instead of iterating for some arbitrrary number of iterations $T$ and taking the final state, instead the population is iterated over all timesteps $t$ covering the timepoints for which there were observations as well as all intervening timesteps. Under this formulation, the population size parameters $\mathbf{n}_{i,t}$ will be affected by the initial population state $\mathbf{n}_{i,0}$. this can either be introduced as an unknown parameter in the model, or set to an arbitrary value, and adding a sufficient 'burn-in' period prior to the first observation such that its value has no little effect on the estimated model parameters.

@@

analysis <- iterate_dynamic_matrix(
  matrix_function = A_t_function, 
  initial_state = n_init,
  niter = timesteps,
  X = X_list,
  beta_F = beta_F,
  beta_S = beta_S,
  alpha_F = alpha_F,
  alpha_S = alpha_S,
  gamma = gamma
)

Lambda <- analysis$all_states
p <- 1 - exp(-Lambda)
distribution(y) <- bernoulli(p)
```
Given: `X_list` a list of length $T$ of design matrices for each of the time points, `n_init` is the initial population state, and `y` is a $N$x$T$ matrix of presence-absence at each timestep and site, and `timesteps` is the value of $T$.

# Transition matrix models

@@

## Multiple species

@@

>>>>

This appendix provides statistical notation and computer code for the R packages greta and greta.dynamics [@goldingGretaDynamicsModelling2019; @goldingGretaSimpleScalable2019] to implement the broad range of DSDMs proposed in the manuscript *Demographic Species Distribution Models* by Nick Golding.

Note that statistical notation is written in the standard way with the first line giving the sampling distribution, and subsequent lines defining the variables on which previous lines depend, however the code uses the reverse order since in greta variables must be defined before they can be used to create other variables.

@@

For ease of reference, I repeat here the equations and interpretation for the basic presence-absence DSDM presented in the main manuscript. It presents the model in its three parts, each of which can be modified independently. Specification of priors for parameters in the model are not discussed here. See [@goldingGretaSimpleScalable2019] and the website [greta-stats.org](https://greta-stats.org) for general guidance on how to build models with greta.

@@

Given vectors `Lambda` and `y` each with length $N$, the number of sites.

## Abundance model

The abundance model relates the transition matrix $\mathbf{A}_i$ of the population at site $i$ to the relative abundance $\Lambda_i$. The intrinsic growth rate $\lambda_i$ of the population is used to compute the theoretical carrying capacity $K_i$ according to the Beverton-Holt model, with competition effect parameter $\alpha$. This theoretical carrying capacity is negative when $\lambda_i < 1$, so it is rectified by the softplus function to always yield a positive abundance $\Lambda_i$, which is close to zero when $\lambda_i < 1$, and approximately equal to $K_i$ when $\lambda_i > 1$.

@@

Given a 3D array `A` with dimensions $N$x$M$x$M$, where $N$ is the number of sites, and $M$ the number of life stages in the population model (dimension of the matrices $\mathbf{A}_i$), and `alpha` is a scalar. When used with its default settings the function `greta.dynamics::iterate_matrix()` returns, among other quantities, a vector of intrinsic growth rates of the matrices in `A` computed by repeated iteration of the matrix

## Population dynamic model

@@

Given an $N$x$K$ matrix `X`, length $K$ vectors `beta_S` and `beta_F`, and a scalar `gamma`, where $N$ is the number of sites and $K$ the number of columns (covariates plus an intercept) of the design matrix.

# Observation models

@@

# Abundance models

@@

iterations <- iterate_dynamic_matrix(
  matrix_function = A_t_function, 
  base_log_F = X %*% beta_F,
  base_logit_S = X %*% beta_S,
  alpha_F = alpha_F,
  alpha_S = alpha_S,
  gamma = gamma
)

Lambda <- iterations$stable_population
```
Given additional scalar parameters `alpha_F` and `alpha_S`.

## Long-term temporal variation

@@

@@

## Short-term temporal variation

Where the time period between observations is sufficiently short that the assumptions above are not met, or when the ecological question being answered requires it, temporal fluctuation in vital rates can instead be modelled by explicitly iterating the population through successive timesteps, at each time step using the corresponding transition matrix, as in the density-dependent vital rate model above. Instead of iterating for some arbitrrary number of iterations $T$ and taking the final state, instead the population is iterated over timesteps $t$ corresponding to time points in the dataset (and all intermediate timesteps). Thus the population at each timestep will be a function not only of the current environmental conditions, but of all previous environmental conditions - capturing the impacts of environmental stochasticity.

Under this formulation, the population size parameters $\mathbf{n}_{i,t}$ will be affected by the initial population state $\mathbf{n}_{i,0}$. this can either be introduced as an unknown parameter in the model, or set to an arbitrary value, and adding a sufficient 'burn-in' period prior to the first observation such that its value has no little effect on the estimated model parameters.

@@

iterations <- iterate_dynamic_matrix(
  matrix_function = A_t_function, 
  initial_state = n_init,
  niter = timesteps,
  X = X_list,
  beta_F = beta_F,
  beta_S = beta_S,
  alpha_F = alpha_F,
  alpha_S = alpha_S,
  gamma = gamma
)

Lambda <- iterations$all_states
p <- 1 - exp(-Lambda)
distribution(y) <- bernoulli(p)
```
Given: `X_list` a list of length $T$ of design matrices for each of the time points, `n_init` is the initial population state, and `y` is an $N$x$T$ matrix of presence-absence at each timestep and site, and `timesteps` is the value of $T$.

# Population dynamic models

@@

<to do, need to package up inteegration code into greta.dynamics?>

## Biotic interactions

<to do, just add multiple life stages>

@@

<to do, two matrix multiplications (mention sparsity) can only be done with density-dependent vital rates>

### 2019-10-16 manuscript/ms.Rmd
tidy up appendix; add regression model bit to main ms

>>>>

# Vital rate regression models

The basic DSDM introduced in the manuscript only considers simple linear models of environmental variables for (appropriately transformed) vital rates. An advantage of specifying these models in a general-purpose statistical modelling software is that these simple models can be replaced with one of any number of statistical models, as appropriate.

For example, theese linear models could be applied to very large numbers of covariates and transformations thereof in combination with regularising priors to perform automatic variable selection, similarly to the popular Maxent SDM software[@phillipsModelingSpeciesDistributions2008]. Or to specify mixed-effects models to account for clustered sampling designs, or use random slopes to share environmental reponse information between species in a biotic interaction DSDM. Instead of linear models, non-linear functions could be introduced with generalisd additive models or Gaussian process models, for which there are existing greta extension packages: greta.gam and greta.gp [@cGretaGamGeneralised2019; @goldingGretaGpGaussian2019]. These can also be used to model spatial and teemporal random effects and account for drivers of species vital rates that are not captured in the measured environmental variables. Prespecified mathematical relationships or covariate values can also be included, enabling incorporation of  spatial and temporal predictions from physiological models.
